---
title: "PGGB and Minigraph summary"
author: "Natalia Zajac"
date: "11/11/2025"
output:
  html_document: 
    highlight: pygments
    theme: sand
    code_folding: true
    toc: true
    toc_float: true
---

# This notebook aims to evaluate how much sequence is retained from each genome in minigraph and summarise statistics from the pangenome (PGGB)

```{r setup, echo=FALSE, include=FALSE, warning=FALSE}
library(vcfR)
library(utils)
library(dplyr)
library(tidyverse)
library(reshape2)
library(cowplot)
library(patchwork)
library(ezRun)
library(ggbeeswarm)
library(data.table)
library(DT)
library(circlize)
library(ComplexHeatmap)
source("functions.R")

cov_minimap2 = NULL
for (i in seq(0,18,1)){
  cov = read.delim(paste0("/home/zajac/SYRI/community", i, "/community", i, ".rPodCre2.1_to_all/coverage.summary.txt"), header = F, sep = "\t")
  cov_minimap2 = rbind(cov_minimap2, cov)
}
cov_minimap2$genome_mapped = str_remove(cov_minimap2$V1, ".fasta.sorted.bam")

cov_minigraph=NULL
for (n in seq(0,18,1)){
  temp=NULL
  for (i in list.files(paste0("../minigraph/community", n), pattern = ".cov.txt", recursive = TRUE)){
    name = str_remove(i, ".fasta.cov.txt")
    df = read.delim(paste0("../minigraph/community", n, "/", i), header = F)
    df$V2 = as.numeric(str_remove(df$V2, "LN:i:"))
    df$V3 = str_remove(df$V3, "SN:Z:")
    df$V4 = str_remove(df$V4, "dc:f:")
    df$V5 = as.numeric(df$V4) * as.numeric(df$V2)
    df$genome_mapped = name
    length_of_mapped_genome=list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = name, recursive = T)[grepl("fai", list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = name, recursive = T))]
    length_of_mapped_genome = read.delim(paste0("../chromosome_sets/community", n, "_unmasked/", length_of_mapped_genome), header = F)
    df$total_length = length_of_mapped_genome$V2
    df = df %>% group_by(genome_mapped) %>% summarise(total_length_mapped = sum(V5), total_length = mean(total_length))
    df$percentage_included = (df$total_length_mapped*100)/df$total_length
    temp = rbind(temp, df)
  }
  cov_minigraph = rbind(cov_minigraph,temp)
}

cov_pggb = NULL
for (n in seq(0,18,1)){
  temp=NULL
  for (i in list.files(paste0("../pggb_allcommunities/community", n, ".s5000"), pattern = "sorted.stats", recursive = TRUE)){
    df = read.delim(paste0("../pggb_allcommunities/community", n, ".s5000/", i), skip = 18)[-27,]
    length_of_mapped_genome=list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T)[grepl("fai", list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T))]
    length_of_mapped_genome = read.delim(paste0("../chromosome_sets/community", n, "_unmasked/", length_of_mapped_genome), header = F)
    df = merge(df,length_of_mapped_genome[,c(1,2)], by.x = "path", by.y = "V1")
    df$percentage = (df$nucleotides*100)/df$V2
    temp = rbind(temp, df[,c(1,8)])
  }
  cov_pggb = rbind(cov_pggb,temp)
}

#Size of pangenome

minigraph_size=NULL
for (n in seq(0,18,1)){
  temp=NULL
  for (i in list.files(paste0("../minigraph/community", n), pattern = "contributing.genome", recursive = TRUE)){
    df = read.delim(paste0("../minigraph/community", n, "/", i), header = F)
    df$V1 = as.numeric(str_remove(df$V1, "LN:i:"))
    df$V2 = str_remove(df$V2, "SN:Z:")
    length_of_mapped_genome=list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T)[grepl("fai", list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T))]
    length_of_mapped_genome = read.delim(paste0("../chromosome_sets/community", n, "_unmasked/", length_of_mapped_genome), header = F)
    df$cretensis_length = length_of_mapped_genome[grepl("rPodCre2.1", length_of_mapped_genome$V1),]$V2
    df = df %>% separate(V2, into = c("species", "hap", "chromosome"), sep = "#")
    temp = rbind(temp, df)
  }
  minigraph_size = rbind(minigraph_size,temp)
}

pggb_size = NULL
for (n in seq(0,18,1)){
  temp=NULL
  for (i in list.files(paste0("../pggb_allcommunities/community", n, ".s5000"), pattern = "sorted.stats", recursive = TRUE)){
    df = read.delim(paste0("../pggb_allcommunities/community", n, ".s5000/", i))[1,]
    length_of_mapped_genome=list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T)[grepl("fai", list.files(paste0("../chromosome_sets/community", n, "_unmasked/"), pattern = "combined", recursive = T))]
    length_of_mapped_genome = read.delim(paste0("../chromosome_sets/community", n, "_unmasked/", length_of_mapped_genome), header = F)
    df$cretensis_length = length_of_mapped_genome[grepl("rPodCre2.1", length_of_mapped_genome$V1),]$V2
    temp = rbind(temp, df)
  }
  pggb_size = rbind(pggb_size,temp)
}
pggb_size$pggb_size = as.numeric(pggb_size$X.length)/pggb_size$cretensis_length

sample_data = read.delim("/groups/mpistaff/Zajac/genomes/sample_data.tsv")
sample_data$clade=factor(sample_data$clade,levels=c("Balkan","Sicilian-Maltese","Western","Siculus","Muralis","Iberian"))
chr_map = read.delim("../chromosome_sets/community.chr.map")
```

## 1. How much of each genome is retained in the analysis

```{r, fig.height=25, fig.width=25, echo=FALSE, warning=FALSE}

merge(merge(cov_minimap2[,c(11,8)] %>% mutate(genome_mapped = str_remove(genome_mapped, ".rev")) , cov_minigraph[,c(1,4)] %>% mutate(genome_mapped = str_remove(genome_mapped, ".rev")), by = "genome_mapped"), cov_pggb, by.x = "genome_mapped" ,by.y = "path") %>% mutate(minimap2 = V8*100, minigraph = if_else(percentage_included > 100, 100,percentage_included) , pggb = percentage) %>% dplyr::select(1,5,6,7) %>% reshape2::melt(id.vars = "genome_mapped") %>% separate(genome_mapped, into = c("species", "hap", "chromosome"), sep = "#") %>% ggplot(aes(paste(species, hap), value, fill = variable)) + geom_col(position = "dodge") + facet_wrap(~chromosome, scales = "free") + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + labs(x = "Genome Haplotype", y = "Percetage of sequence retained in the analysis")

```

## 2. How much does each chromosome increase if all sequence included

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

merge(minigraph_size %>%  group_by(chromosome) %>% summarise(minigraph_size = sum(V1), cretensis_length = mean(cretensis_length)), pggb_size, by = "cretensis_length") %>% mutate(minigraph_size = minigraph_size/cretensis_length) %>% dplyr::select(2,3,9) %>% reshape2::melt(id.vars = "chromosome") %>% ggplot(aes(chromosome, value, fill = variable)) + geom_col(position = "dodge") + labs(y = "Increase in size in comparison to P.cretensis genome")

```

## 3. Does pggb have variants in regions unmapped in minimap2

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

cov_matters = NULL

for (i in seq(0,18,1)){
  sp=NULL
  for (x in unique(minigraph_size$species)){
    depth_files = list.files(paste0("../SYRI/community", i, "/community", i, ".rPodCre2.1_to_all/"), pattern = x)[grepl("depth", list.files(paste0("../SYRI/community", i, "/community", i, ".rPodCre2.1_to_all/"), pattern = x))]
    dp = NULL
    for (l in depth_files){
      dpt = read.delim(paste0("../SYRI/community", i, "/community", i, ".rPodCre2.1_to_all/", l), header = F)
      dp = rbind(dpt, dp)
    }
    colnames(dp) = c("chrom", "pos", "depth")
    dp = dp %>% group_by(chrom, pos) %>% summarise(depth = max(as.numeric(depth))) %>% arrange(as.numeric(pos))
    dp$group <- cumsum(c(TRUE, dp$depth[-1] != dp$depth[-nrow(dp)]))
    dp = dp %>% group_by(chrom, group, depth) %>% summarise("start" = min(pos), "end" = max(pos))
    dp = dp[,c(1,3,4,5)]
    dp = makeGRangesFromDataFrame(dp, seqnames.field = "chrom", start.field = "start", end.field = "end", keep.extra.columns = T)
    pggb_SNP=read.vcfR(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".", x, ".pggb.waved.norm.SNP.vcf"))
    pggb_SNP=data.frame(pggb_SNP@fix)
    pggb_SV=read.vcfR(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".", x, ".pggb.waved.norm.SV.vcf"))
    pggb_SV=data.frame(pggb_SV@fix)
    pggb_MNP=read.vcfR(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".", x, ".pggb.waved.norm.MNP.vcf"))
    pggb_MNP=data.frame(pggb_MNP@fix)
    pggb_all = rbind(pggb_SNP[,c(1,2)] %>% mutate(cat = "SNP"), pggb_SV[,c(1,2)] %>% mutate(cat = "MNP"), pggb_MNP[,c(1,2)] %>% mutate(cat = "SV"))
    pggb_all = makeGRangesFromDataFrame(pggb_all, seqnames.field = "CHROM", start.field = "POS", end.field = "POS", keep.extra.columns = T)
    hits = findOverlaps(pggb_all, dp)
    query_hits   <- pggb_all[queryHits(hits)]
    subject_hits <- dp[subjectHits(hits)]
    overlaps <- pintersect(query_hits, subject_hits)
    result <- data.frame(
    query_seqname   = as.character(seqnames(query_hits)),
    query_start     = start(query_hits),
    query_end       = end(query_hits),
    query_cat       = query_hits$cat,
    subject_seqname = as.character(seqnames(subject_hits)),
    subject_start   = start(subject_hits),
    subject_depth   = subject_hits$depth,
    subject_end     = end(subject_hits)) %>% unique()
    result = result %>% group_by(query_cat,subject_depth, query_seqname) %>% summarise(n_distinct(query_start)) %>% mutate(subject_depth = if_else(subject_depth != "0", "1+", "0")) %>% group_by(query_seqname, query_cat, subject_depth) %>% summarise(variant_nb = sum(`n_distinct(query_start)`)) %>% mutate(species = x) 
    sp = rbind(result, sp)
  }
  cov_matters = rbind(sp, cov_matters)
}

cov_matters %>% ggplot(aes(species, variant_nb, fill = query_cat)) + geom_col(position = "dodge") + facet_wrap(~subject_depth) + scale_y_log10() + scale_fill_manual(values = c("cornflowerblue", "gold3", "darkgreen")) + theme_bw() + labs(x = "species", y = "Number of variants",  fill = "variant type")

```

## 4. What is core/dispensable/private

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

growth_curve = NULL
for (i in seq(0,18,1)){
  comm = paste0("community",i)
  spstats1 = read.delim(paste0("../pggb_allcommunities/community",i, ".s5000/community",i, ".sorted.stats"), sep = "\t", header = F,skip = 2)[1:15,]
  colnames(spstats1) = c("sample", "core", "private", "dispensable")
  if (is.null(growth_curve)){
    growth_curve = spstats1
  } else {
    growth_curve = merge(growth_curve,spstats1, by = "sample")
    growth_curve = growth_curve %>% mutate(across(all_of(colnames(growth_curve)[-1]), ~ as.numeric(.)))
    growth_curve$core = growth_curve$core.x + growth_curve$core.y
    growth_curve$private = growth_curve$private.x + growth_curve$private.y
    growth_curve$dispensable = growth_curve$dispensable.x + growth_curve$dispensable.y
    growth_curve = growth_curve[,c(1,8,9,10)]
  }
}
growth_curve$sample = sample_data[match(growth_curve$sample, sample_data$MyName),]$Organism.Name
growth_curve$clade = sample_data[match(growth_curve$sample, sample_data$Organism.Name),]$clade
species = growth_curve %>% mutate(total = as.numeric(private) + as.numeric(dispensable)) %>% arrange(as.numeric(total)) %>% dplyr::select(sample) %>% as.list()

p1 = growth_curve %>% reshape2::melt(id.vars = c("sample", "clade")) %>% ggplot(aes(sample, as.numeric(value)/1000000000 , fill = factor(variable, levels = c("private", "dispensable", "core")))) + geom_col(position = "stack") + labs(x = "species", y = "Sequence length (Gbp)", fill = "PG Sequence Class") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)) + scale_fill_manual(values = c("#cc79a7", "orange","#0072b2")) + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust =1, face = "italic", size = 9), legend.position = "top", legend.text = element_text(size = 12))

pdf("../Figures/PG_size_perSpecies.pdf", width = 8, height = 4)
p1 + facet_grid(~clade, space = "free", scales = "free")
dev.off()



##Alternative
growth_curve$total = growth_curve$private + growth_curve$dispensable + growth_curve$core
growth_curve = growth_curve %>% mutate(dispensable = dispensable*100/total, core= core*100/total, private= private*100/total) %>% dplyr::select(1:5) %>% reshape2::melt(id.vars = c("sample", "clade")) 
growth_curve$variable = factor(growth_curve$variable, levels = c("core", "dispensable", "private"))
growth_curve$sample = factor(growth_curve$sample, levels = c("Podarcis gaigeae", "Podarcis melisellensis", "Podarcis erhardii", "Podarcis cretensis", "Podarcis filfolensis", "Podarcis raffonei", "Podarcis pityusensis", "Podarcis lilfordi", "Podarcis tiliguerta", "Podarcis bocagei", "Podarcis vaucheri","Podarcis liolepis", "Podarcis muralis nigriv.", "Podarcis muralis", "Podarcis siculus"))
growth_curve$clade = factor(growth_curve$clade, levels = rev(c("Siculus", "Muralis", "Iberian", "Western", "Sicilian-Maltese", "Balkan")))

pdf("../Figures/PG_size_perSpecies.pdf", width = 5, height = 8)
ggplot(growth_curve, aes(value,sample, fill = variable)) + geom_col() + facet_grid(clade~., scales = "free", space = "free") + scale_fill_manual(values = c("#0072b2", "orange","#cc79a7")) + theme_bw() + theme(axis.text.y = element_text(face = "italic"), legend.position = "top") + labs(x = "Genome length (%)", y = "", fill = "Class")
dev.off()
```

## 4. What is core/dispensable/private but per chromosome

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

growth_curve = NULL
for (i in seq(0,18,1)){
  comm = paste0("community",i)
  spstats1 = read.delim(paste0("../pggb_allcommunities/community",i, ".s5000/community",i, ".sorted.stats"), sep = "\t", header = F,skip = 2)[1:15,]
  colnames(spstats1) = c("sample", "core", "private", "dispensable")
  spstats1$community = i
  growth_curve = rbind(spstats1,growth_curve)
}
pdf("../Figures/PG_size_classes_perCHR.pdf", width = 5, height = 8)
merge(growth_curve, chr_map, by = "community") %>% reshape2::melt(id.vars = c("sample","community", "chromosome")) %>% arrange() %>% ggplot(aes(as.numeric(value), factor(chromosome, levels = c(as.character(seq(1,18,1)), "Z")), fill = factor(variable, levels = c("core", "dispensable", "private")))) + geom_col(position = "fill") + theme_bw()  + scale_fill_manual(values = c("darkcyan", "darkblue","pink3")) + theme(legend.position = "top", text = element_text(size =18)) + labs(y = "Chromosome", x = "Proportion", fill = "Class")
dev.off()
```

## 5. Compression ratio per chromsome

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

comp_ratio = NULL
for (i in seq(0,18,1)){
  comm = paste0("community",i)
  size = read.delim(paste0("../pggb_allcommunities/community",i, ".s5000/community",i, ".sorted.stats"), sep = "\t", header = T)[1,]$X.length
  chromosome = unique(sapply(str_split(read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)$V1, "#"), .subset, 3))
  mean_chr_size = read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)[,c(1,2)] %>% mutate(V1 = sapply(str_split(V1, "#"), .subset, 1)) %>% group_by(V1) %>% summarise(V2 = mean(V2)) %>% mutate(V3 = size) %>% mutate(V4 = chromosome)
  comp_ratio = rbind(comp_ratio, mean_chr_size)
}

pdf("../Figures/PG_compressionR_perChr.pdf", width = 4, height = 5)
comp_ratio$V4 = factor(comp_ratio$V4, levels = c(seq(1,18,1), "Z"))
ggplot(comp_ratio, aes(as.numeric(V3)/as.numeric(V2), V4 )) + geom_boxplot(fill = "skyblue") + theme_linedraw()  + labs(x= "Size of the graph/Size of the chromosome", y = "Chromosome")
dev.off()

```

###The two above combined

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
growth_curve = merge(growth_curve, chr_map, by = "community") 
growth_curve$size = as.numeric(growth_curve$private) + as.numeric(growth_curve$core) + as.numeric(growth_curve$dispensable)
growth_curve = growth_curve %>% mutate(core = as.numeric(core)*100/size, private = as.numeric(private)*100/size, dispensable = as.numeric(dispensable)*100/size) %>% reshape2::melt(id.vars = c("sample","community", "chromosome", "size")) %>% group_by(chromosome, variable) %>% dplyr::summarise(mean(as.numeric(value)))
growth_curve$variable = factor(growth_curve$variable, levels = c("core", "dispensable", "private"))

pdf("../Figures/PG_size_classes_perCHR_compressionratio.pdf", height = 6, width = 8)
merge(growth_curve, comp_ratio, by.x = "chromosome", by.y = "V4") %>% ggplot(aes(as.numeric(V3)/as.numeric(V2), `mean(as.numeric(value))`, colour = variable)) + geom_point(size = 3) + geom_label_repel(aes(label = chromosome)) + scale_colour_manual(values = c("#0072b2", "orange","#cc79a7")) + theme_bw() + labs(x = "Compression ratio (pangenome size/mean species length)", y = "% of pangenome", colour = "Class")
dev.off()
```

## 5A. Compression ratio per chromsome only first haplotypes

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

comp_ratio = NULL
for (i in seq(0,18,1)[c(-4,-10)]){
  comm = paste0("community",i)
  size = read.delim(paste0("../pggb_allcommunities/community",i, ".s5000/community",i, ".sorted.single_haps.stats"), sep = "\t", header = T)[1,]$X.length
  chromosome = unique(sapply(str_split(read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)$V1, "#"), .subset, 3))
  mean_chr_size = read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)[,c(1,2)] %>% mutate(V1 = sapply(str_split(V1, "#"), .subset, 1)) %>% group_by(V1) %>% summarise(V2 = mean(V2)) %>% mutate(V3 = size) %>% mutate(V4 = chromosome)
  comp_ratio = rbind(comp_ratio, mean_chr_size)
}
i=9
comm = paste0("community9")
size = read.delim(paste0("../pggb_allcommunities/community",i, ".s5000/community",i, ".sorted.stats"), sep = "\t", header = T)[1,]$X.length
chromosome = unique(sapply(str_split(read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)$V1, "#"), .subset, 3))
mean_chr_size = read.delim(paste0("../chromosome_sets/", comm, "_unmasked/", comm, ".combined.fasta.fai"), header = F)[,c(1,2)] %>% mutate(V1 = sapply(str_split(V1, "#"), .subset, 1)) %>% group_by(V1) %>% summarise(V2 = mean(V2)) %>% mutate(V3 = size) %>% mutate(V4 = chromosome)
comp_ratio = rbind(comp_ratio, mean_chr_size)


comp_ratio %>% mutate(chromosome_type = if_else(V4 == "Z", "Z", "autosome")) %>% data.frame() %>% ggplot( aes(as.numeric(V3)/as.numeric(V2), chromosome_type )) + geom_boxplot(fill = "skyblue") + theme_linedraw()  + labs(x= "Size of the graph/Size of the chromosome", y = "Chromosome")

comp_ratio$V4 = factor(comp_ratio$V4, levels = c(seq(1,18,1), "Z"))
ggplot(comp_ratio, aes(as.numeric(V3)/as.numeric(V2), V4 )) + geom_boxplot(fill = "skyblue") + theme_linedraw()  + labs(x= "Size of the graph/Size of the chromosome", y = "Chromosome")
```

## 5B. Dendorgram per chromosome

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
all_dendros = NULL
for (i in seq(0,18,1)){
  chr = paste0("chr", chr_map[chr_map$community == i,]$chromosome)
  df = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".smooth.final.dist.tsv"))
  jaccard_dist_df <- df %>%
    arrange(group.a, group.b) %>%
    select(group.a, group.b, jaccard.distance) %>%
    pivot_wider(names_from = group.b, values_from = jaccard.distance) %>%
    column_to_rownames(var = "group.a")

# Clustering
  jaccard_hc <- as.dist(jaccard_dist_df) %>% hclust()
  tree = as.phylo(jaccard_hc)
  all_dendros[[chr]] = tree
}
class(all_dendros) <- "multiPhylo"
sample_map = c(setNames(paste0(sample_data$Organism.Name, "#1"),paste0(sample_data$MyName, "#1")),setNames(paste0(sample_data$Organism.Name, "#2"),paste0(sample_data$MyName, "#2")))
for (i in seq(1,19,1)){ all_dendros[[i]]$tip.label = unname(unlist(sample_map[all_dendros[[i]]$tip.label]))}


pdf("../Figures/Densitree.pdf", width = 8, height = 6)
densiTree(all_dendros, type = "phylogram", scaleX = TRUE, alpha = 0.1)
dev.off()
```

## 6. Node numbers vs node lengths per chromosome in core/dispensable/private

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
nodes_lengths = NULL
for (i in seq(0,18,1)){
  private = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".sorted.allSpecies.nodes.private"), header = F)
  private$cat = "private"
  dispensable = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".sorted.allSpecies.nodes.dispensable"), header = F)
  dispensable$cat = "dispensable"
  core = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".sorted.allSpecies.nodes.core"), header = F)
  core$cat = "core"
  all = rbind(core, private, dispensable)
  length_nodes = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/community", i, ".sorted.nodes.fasta.len"), header = F)
  all = merge(all, length_nodes, by = "V1")
  all = all %>% group_by(V2,cat) %>% summarise(nodes = n_distinct(V1))
  all$comm = i
  nodes_lengths = rbind(all, nodes_lengths)
}
pdf("../Figures/Node_length.pdf", width = 6, height = 4)
nodes_lengths %>% ggplot(aes(as.numeric(V2), as.numeric(nodes), colour = cat)) + geom_point() + scale_x_log10() + scale_y_log10() + labs(x = "Node length (bp)", y = "Frequency") + theme_bw() + scale_colour_manual(values = c("darkcyan", "darkblue","pink3"))
dev.off()
```


## 7. Graph-based pangenome modelling selfmade

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
all_odgi=NULL
for (i in seq(0,18,1)){
  odgi = read.delim(paste0("../pggb_allcommunities/community", i, ".s5000/growth.curve.csv"), sep = ",")
  odgi$com = i
  all_odgi = rbind(odgi, all_odgi)
}

pdf("../Figures/PG_GrowthCurve.pdf", width = 4, height = 4)
rbind(all_odgi %>% group_by(class,nb_genomes,com) %>% dplyr::summarise(mean(value)) %>% group_by(class,nb_genomes) %>% dplyr::summarise(sum(`mean(value)`)) %>% mutate(size = `sum(\`mean(value)\`)`/1000000000), all_odgi %>% group_by(class,nb_genomes,com) %>% dplyr::summarise(mean(value))  %>% group_by(nb_genomes) %>% dplyr::summarise(sum(`mean(value)`)) %>% mutate(size = `sum(\`mean(value)\`)`/1000000000)  %>% mutate(class = "total")) %>% ggplot(aes(nb_genomes,size, group = class, colour = class)) + geom_point() +  geom_smooth() + theme_bw() + labs(x= "nth genome", y = "Sequence length (Gbp)") + scale_colour_manual(values = c("#0072b2", "orange", "#cc79a7","black")) + theme(legend.position = "top")
dev.off()

x = all_odgi %>%
    group_by(nb_genomes, com, class) %>%
    mutate(replicate = row_number()) %>%
    ungroup()

n_comm <- x %>% distinct(com) %>% nrow()

x = x %>% group_by(nb_genomes, class, replicate) %>%
  filter(n_distinct(com) == n_comm) %>%
  ungroup()

pdf("../Figures/PG_GrowthCurve.pdf", width = 6, height = 4)
rbind(rbind(x %>% group_by(nb_genomes, class, replicate) %>% summarise(size = sum(value)/1000000000), data.frame(nb_genomes=c(1,1), class = c("private", "dispensable"), replicate=c(1,1), size = c(0,0))), x %>% group_by(nb_genomes,replicate) %>% summarise(size = sum(value)/1000000000) %>% mutate(class = "total")) %>% ggplot(aes(nb_genomes,size, group = class, colour = class)) + geom_point() +  geom_smooth() + theme_bw() + labs(x= "nth genome", y = "Sequence length (Gbp)") + scale_colour_manual(values = c("darkcyan", "darkblue", "pink3", "black")) + theme(legend.position = "top")
dev.off()


```

## 8. Shared node space between and within haps and clades

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

shared_node_space_haps=NULL
for (i in c(seq(0,8,1), seq(10,18,1))){
  df = read.delim(paste0("/home/zajac/pggb_allcommunities/community", i, ".s5000/odgi.sharednodespace.haps.txt"))
  shared_node_space_haps = rbind(shared_node_space_haps,df)
}
shared_node_space_haps = shared_node_space_haps %>% mutate(group.name = sapply(str_split(query, "#"), .subset, 1)) %>% mutate(cat = "Within haplotypes")
shared_node_space_haps$group.name = factor(shared_node_space_haps$group.name, levels = c("rPodBoc1", "rPodLio1","rPodVau1", "rPodMur119", "rPodSic1", "rPodErh1","rPodGai1","rPodMel1","rPodFil1","rPodTil1","rPodPit1"))



shared_node_space_clades=NULL
for (i in seq(0,18,1)){
  df = read.delim(paste0("/home/zajac/pggb_allcommunities/community", i, ".s5000/odgi.sharednodespace.clades.txt"))
  shared_node_space_clades = rbind(shared_node_space_clades,df)
}
shared_node_space_clades = shared_node_space_clades %>% mutate(group.name = case_when(group.name == "group1" ~ "Muralis", group.name == "group2" ~ "Iberian", group.name == "group3" ~ "Balkan", group.name == "group4" ~ "Sicilian-Maltese", group.name == "group5" ~ "Western Islands", group.name == "group6" ~ "Siculus")) %>% mutate(cat = "Within 6 major clades")
shared_node_space_clades$group.name = factor(shared_node_space_clades$group.name, levels = c("Iberian","Muralis","Siculus", "Balkan", "Sicilian-Maltese","Western Islands"))

shared_node_space_lin=NULL
for (i in seq(0,18,1)){
  df = read.delim(paste0("/home/zajac/pggb_allcommunities/community", i, ".s5000/odgi.sharednodespace.lineages.txt"))
  shared_node_space_lin = rbind(shared_node_space_lin,df)
}
shared_node_space_lin = shared_node_space_lin %>% mutate(group.name = case_when(group.name == "group1" ~ "lineage1", group.name == "group2" ~ "lineage2")) %>% mutate(cat = "")
shared_node_space_lin$group.name = factor(shared_node_space_lin$group.name, levels = c("lineage1", "lineage2"))

shared = rbind(shared_node_space_haps,shared_node_space_clades,shared_node_space_lin)
shared = shared %>% mutate(col = case_when(group.name %in% c("lineage1", "Iberian","Muralis","Siculus", "rPodBoc1", "rPodLio1","rPodVau1", "rPodMur119", "rPodSic1") ~ "Iberian/Muralis/Siculus", group.name %in% c("rPodErh1","rPodGai1","rPodMel1","rPodFil1","rPodTil1","rPodPit1", "Balkan", "Sicilian-Maltese","Western Islands", "lineage2") ~ "Balkan/Sicilian-Maltese/Western-Islands"))
shared$col = factor(shared$col, levels = c("Iberian/Muralis/Siculus", "Balkan/Sicilian-Maltese/Western-Islands"))

ggplot(shared, aes(group.name, overlap.frac, fill = col)) + geom_violin() + facet_grid(~cat, space = "free", scales = "free") + theme_bw() + theme( axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position = "top", text = element_text(size = 25)) + labs(x = "", y = "Shared node space", fill = "")
```

## 9. Shared node space but in the form of circularization

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

df = shared_node_space_haps %>% group_by(`group.name`) %>% dplyr::summarise(mean(overlap.frac)) %>% mutate(group.name1 = paste0(group.name, "#1"), group.name2 = paste0(group.name, "#2"))
colnames(df) = c("species", "shared_percentage", "haplotype1", "haplotype2")
metadata = sample_data[,c(16,17)] %>% dplyr::rename("species" = "MyName")
df <- df %>% left_join(metadata, by = "species") %>% arrange(clade, species)
df$shared_percentage = df$shared_percentage * 100
sectors <- c(rbind(df$haplotype1, df$haplotype2))

# Create clade mapping for each sector
sector_clade <- rep(df$clade, each = 2)
names(sector_clade) <- sectors

# Set sector sizes
sector_sizes <- rep(100, length(sectors))
names(sector_sizes) <- sectors

# Create link data - connecting haplotype pairs
links <- data.frame(
    from = df$haplotype1,
    to = df$haplotype2,
    value = df$shared_percentage
)

# Color gradient based on shared percentage
col_fun <- colorRamp2(c(75, 95), 
                      c("#3498db", "#e74c3c"))

# Clade colors for the outer track
clade_colors <- c("purple","palegreen3",  "blue3","yellow", "green", "magenta2")

pdf("../Figures/cirlize_haplotypes_shared_space.pdf", width = 12, height = 8)
circos.clear()
circos.par(
  start.degree = 90,
  gap.degree = c(rep(1, length(sectors) - 1), 8),  # Larger gap at the end
  track.margin = c(0.01, 0.01),
  cell.padding = c(0.02, 0, 0.02, 0)
)

# Initialize with sectors
circos.initialize(factors = sectors, xlim = cbind(rep(0, length(sectors)), sector_sizes))

# Track 1: Clade labels (outermost)
circos.track(
  factors = sectors,
  ylim = c(0, 1),
  track.height = 0.08,
  bg.border = NA,
  panel.fun = function(x, y) {
    sector.name = get.cell.meta.data("sector.index")
    clade = sector_clade[sector.name]
    
    # Add clade color background
    circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                col = clade_colors[clade], border = "white", lwd = 0.5)
  }
)

# Add clade labels (only once per clade group)
clade_positions <-df %>%
    group_by(clade) %>%
    summarise(
        start_sector = dplyr::first(haplotype1),
        end_sector = dplyr::last(haplotype2),
        .groups = 'drop'
    )

for(i in 1:nrow(clade_positions)) {
  # Get the start and end positions for this clade
  start_sector <- clade_positions$start_sector[i]
  end_sector <- clade_positions$end_sector[i]
  
  # Calculate middle angle
  start_deg <- get.cell.meta.data("cell.start.degree", sector.index = start_sector, track.index = 1)
  end_deg <- get.cell.meta.data("cell.end.degree", sector.index = end_sector, track.index = 1)
  mid_deg <- (start_deg + end_deg) / 2
  
}

# Track 2: Haplotype labels and scale
circos.track(
  factors = sectors,
  ylim = c(0, 1),
  track.height = 0.15,
  bg.border = NA,
  panel.fun = function(x, y) {
    sector.name = get.cell.meta.data("sector.index")
    
    # Add sector background color
    if(grepl("#1", sector.name)) {
      circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                  col = "#34495e", border = "white", lwd = 0.5)
    } else {
      circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                  col = "#2c3e50", border = "white", lwd = 0.5)
    }
    
    # Add axis with scale
    circos.axis(h = "top", 
                major.at = seq(0, 100, by = 20),
                labels.cex = 0.4,
                labels.niceFacing = TRUE,
                minor.ticks = 1)
    
    # Add sector labels (haplotype names)
    circos.text(
      x = get.cell.meta.data("xlim")[2]/2,
      y = 0.5,
      labels = gsub("_", " ", sector.name),
      cex = 0.5,
      facing = "bending.inside",
      niceFacing = TRUE,
      col = "white",
      font = 2
    )
  }
)

# Add links between haplotype pairs
for(i in 1:nrow(links)) {
  # Get color for this link
  link_col <- col_fun(links$value[i])
  
  circos.link(
    sector.index1 = links$from[i],
    point1 = c(0, links$value[i]),  # Width represents shared space
    sector.index2 = links$to[i],
    point2 = c(0, links$value[i]),
    col = link_col,
    border = NA,
    h.ratio = 0.6  # Height of the link curve
  )
}

# Add legend for color scale
lgd = Legend(
  at = seq(floor(75), ceiling(95), length.out = 5),
  col_fun = col_fun,
  title = "Shared Node\nSpace (%)",
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 8),
  legend_height = unit(3, "cm"),
  grid_width = unit(0.5, "cm")
)

# Draw legend
pushViewport(viewport(x = 0.85, y = 0.5))
grid.draw(lgd)
upViewport()

circos.clear()
dev.off()
# Print summary statistics by clade
cat("\nSummary of Shared Node Space by Clade:\n")
df %>%
  group_by(clade) %>%
  summarise(
    n_species = n(),
    mean_shared = round(mean(shared_percentage), 2),
    min_shared = round(min(shared_percentage), 2),
    max_shared = round(max(shared_percentage), 2),
    .groups = 'drop'
  ) %>%
  print()

pdf("../Figures/cirlize_haplotypes_shared_space_barplot.pdf", height = 5,width = 10)
merge(df, sample_data[,c(3,17)], by.x = "species", by.y = "MyName") %>% ggplot(aes(shared_percentage, factor(Organism.Name, levels = merge(df, sample_data[,c(3,17)], by.x = "species", by.y = "MyName") %>% arrange(clade) %>% pull(Organism.Name)), fill = clade)) + geom_col() + theme_bw() + scale_fill_manual(values = c( "purple","palegreen3","blue3","yellow","green", "magenta2")) + theme( axis.text.y = element_text(size = 15,face = "italic"), axis.text.x = element_text(size = 15)) + labs(x = "Shared node space (%)", y = "")
dev.off()
```


```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

df = shared_node_space_clades %>% mutate(chr = sapply(str_split(query,"#"), .subset, 3)) %>% mutate(query = paste0(sapply(str_split(query,"#"), .subset, 1), "#", sapply(str_split(query,"#"), .subset, 2))) %>% mutate(target = paste0(sapply(str_split(target,"#"), .subset, 1), "#", sapply(str_split(target,"#"), .subset, 2)))
df = df %>% group_by(group.name, query,target) %>% dplyr::summarise(overlap.frac = mean(overlap.frac) * 100)
colnames(df) = c("clade", "haplotype1", "haplotype2", "shared_percentage")
df = df[sapply(str_split(df$haplotype1,"#"), .subset, 1) != sapply(str_split(df$haplotype2,"#"), .subset, 1),]

species_haplotypes = data.frame(genome= unique(c(df$haplotype1, df$haplotype2))) %>% mutate(species = sapply(str_split(genome,"#"), .subset, 1)) %>% group_by(species) %>% summarise(n_haps = n_distinct(genome))
all_haplotypes = unique(c(df$haplotype1, df$haplotype2))

df$species1 = sapply(str_split(df$haplotype1,"#"), .subset, 1)
df$species2 = sapply(str_split(df$haplotype2,"#"), .subset, 1)

species_info <- species_haplotypes %>%
    left_join(metadata, by = "species") %>%
    arrange(clade, species)
species_info = unique(species_info)

sectors <- unlist(lapply(1:nrow(species_info), function(i) {
    sp <- species_info$species[i]
    n <- species_info$n_haps[i]
    if(n == 1) {
        paste0(sp, "#1")
    } else {
        c(paste0(sp, "#1"), paste0(sp, "#2"))
    }
}))
sectors = unique(sectors)

sector_info <- data.frame(
    sector = sectors,
    species = rep(species_info$species, species_info$n_haps),
    clade = rep(species_info$clade, species_info$n_haps),
    stringsAsFactors = FALSE
)

sector_sizes <- rep(100, length(sectors))
names(sector_sizes) <- sectors

# Color gradient based on shared percentage
col_fun <- colorRamp2(c(75, 95), 
                      c("#3498db", "#e74c3c"))

clade_colors <- c("purple","palegreen3",  "blue3","yellow", "green", "magenta2")

gaps <- sapply(1:(length(sectors)-1), function(i) {
    current_sp <- sector_info$species[i]
    next_sp <- sector_info$species[i+1]
    current_clade <- sector_info$clade[i]
    next_clade <- sector_info$clade[i+1]
    
    if(current_clade != next_clade) {
        return(8)  # Large gap between clades
    } else if(current_sp != next_sp) {
        return(3)  # Medium gap between species
    } else {
        return(0.5)  # Small gap between haplotypes of same species
    }
})
gaps <- c(gaps, 10) 

pdf("../Figures/cirlize_clades_shared_space.pdf", width = 12, height = 8)
circos.clear()
circos.par(
  start.degree = 90,
  gap.degree = gaps,
  track.margin = c(0.01, 0.01),
  cell.padding = c(0.02, 0, 0.02, 0)
)
circos.initialize(factors = sectors, xlim = cbind(rep(0, length(sectors)), sector_sizes))
circos.track(
  factors = sectors,
  ylim = c(0, 1),
  track.height = 0.08,
  bg.border = NA,
  panel.fun = function(x, y) {
    sector.name = get.cell.meta.data("sector.index")
    clade = sector_info$clade[sector_info$sector == sector.name]
    
    # Add clade color background
    circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                col = clade_colors[clade], border = "white", lwd = 0.5)
  }
)
clade_positions <- sector_info %>%
    group_by(clade) %>%
    summarise(
        start_sector = dplyr::first(sector),
        end_sector = dplyr::last(sector),
        .groups = 'drop'
    )

# Track 2: Haplotype labels and scale
circos.track(
    factors = sectors,
    ylim = c(0, 1),
    track.height = 0.15,
    bg.border = NA,
    panel.fun = function(x, y) {
        sector.name = get.cell.meta.data("sector.index")
        
        # Add sector background color
        if(grepl("#1", sector.name)) {
            circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                        col = "#34495e", border = "white", lwd = 0.5)
        } else {
            circos.rect(0, 0, get.cell.meta.data("xlim")[2], 1, 
                        col = "#2c3e50", border = "white", lwd = 0.5)
        }
        
        # Add axis with scale
        circos.axis(h = "top", 
                    major.at = seq(0, 100, by = 20),
                    labels.cex = 0.4,
                    labels.niceFacing = TRUE,
                    minor.ticks = 1)
        
        # Add sector labels (haplotype names)
        circos.text(
            x = get.cell.meta.data("xlim")[2]/2,
            y = 0.5,
            labels = gsub("_", " ", sector.name),
            cex = 0.5,
            facing = "bending.inside",
            niceFacing = TRUE,
            col = "white",
            font = 2
        )
    }
)

# Add links between haplotypes
for(i in 1:nrow(df)) {
    # Get color for this link
    link_col <- col_fun(df$shared_percentage[i])
    
    circos.link(
        sector.index1 = df$haplotype1[i],
        point1 = c(0, df$shared_percentage[i]),
        sector.index2 = df$haplotype2[i],
        point2 = c(0, df$shared_percentage[i]),
        col = link_col,
        border = NA,
        h.ratio = 0.6
    )
}

# Add legend for color scale
lgd = Legend(
    at = seq(floor(75), ceiling(95), length.out = 5),
    col_fun = col_fun,
    title = "Shared Node\nSpace (%)",
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 8),
    legend_height = unit(3, "cm"),
    grid_width = unit(0.5, "cm")
)

# Draw legend
pushViewport(viewport(x = 0.85, y = 0.5))
grid.draw(lgd)
upViewport()
dev.off()
# Print summary statistics
cat("\nSpecies Haplotype Summary:\n")
species_info %>%
  group_by(clade) %>%
  summarise(
    n_species = n(),
    species_with_1_hap = sum(n_haps == 1),
    species_with_2_haps = sum(n_haps == 2),
    .groups = 'drop'
  ) %>%
  print()

cat("\nInter-species Shared Node Space Summary:\n")
print(summary(df$shared_percentage))

pdf("../Figures/cirlize_clades_shared_space_boxplot.pdf", height = 6,width = 8)
df %>% group_by(species1,species2, clade) %>% summarise(shared_percentage = mean(shared_percentage)) %>% ggplot(aes(shared_percentage, clade, fill = clade)) + geom_boxplot() + theme_bw() + scale_fill_manual(values = c("magenta2", "green", "purple","palegreen3",  "blue3")) + theme(legend.position = "", axis.text.y = element_text(size = 15), axis.text.x = element_text(size = 15)) + labs(x = "Shared node space (%)", y = "")
dev.off()
```


## 10. Phylogenetically congruent dispensable nodes - visualising 04_parsimonious.dispnodes.R

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

all_nonp = NULL
for (i in list.files("../pggb_allcommunities/", pattern = "nonparsimonious*", recursive = T)){
  df = read.delim(paste0("../pggb_allcommunities/",i))
  if (is.null(all_nonp)){
    all_nonp = df
  } else {
   all_nonp = merge(all_nonp, df, by = "genome", all = TRUE) 
  }
}
all_nonp = data.frame(nonp_genome = rowSums(all_nonp %>% column_to_rownames("genome")))
genom_size = NULL
for (i in list.files("/groups/mpistaff/Zajac/genomes/ncbi_dataset/data/ALL_corrected_unmasked/", pattern = "*.fai")) {
  name = str_remove(i, ".all.fasta.fai")
  value = as.character(tail(read_delim(paste0("/groups/mpistaff/Zajac/genomes/ncbi_dataset/data/ALL_corrected_unmasked/", i)), n = 1)[,3])
  df = data.frame("hap" = name, "size" = value)
  genom_size = rbind(genom_size, df)
}
all_nonp = merge(all_nonp %>% rownames_to_column("genome"), genom_size %>% mutate(genome = sapply(str_split(hap, "#"), .subset, 1)) %>% group_by(genome) %>% summarise(size = mean(as.numeric(size))), by = "genome") %>% mutate(perc = (nonp_genome * 100)/size)
sample_data = read.delim("/groups/mpistaff/Zajac/genomes/sample_data.tsv")
all_nonp = merge(all_nonp, sample_data[,c(3,17)], by.x = "genome", by.y = "MyName") %>% unique()
all_nonp$Organism.Name = factor(all_nonp$Organism.Name, levels = c("Podarcis cretensis", "Podarcis erhardii", "Podarcis gaigeae", "Podarcis melisellensis","Podarcis filfolensis", "Podarcis raffonei" , "Podarcis lilfordi", "Podarcis pityusensis", "Podarcis tiliguerta", "Podarcis siculus",  "Podarcis muralis", "Podarcis muralis nigriv.", "Podarcis liolepis", "Podarcis bocagei" ,"Podarcis vaucheri"))

pdf("../Figures/Phylogenetically_disc_disp_genome.pdf", height = 6, width = 6)
all_nonp %>% ggplot(aes(perc, Organism.Name)) + geom_col(colour = "black", fill = "black") + theme_bw() + theme(text = element_text(size = 20), axis.text.y = element_text(face = "italic"), title = element_text(size = 9))  + labs(x = "Percentage of genome size", y = "Species") + ggtitle("Phylogenetically discordant dispensable genome")
dev.off()

pdf("../Figures/disc_disp_genome_vs_priv.pdf", height = 6, width = 9)
merge(growth_curve[,c(1,3)], all_nonp, by.x = "sample", by.y = "Organism.Name") %>% mutate(perc_priv = (private * 100)/size) %>% ggplot(aes(perc, perc_priv)) + geom_point(size = 3) + theme_bw() + labs(y = "Percentage of private genome",x = "Percentage of phylogenetically discordant dispensable genome") + theme(text = element_text(size = 20)) + geom_smooth(method=lm , color="red", se=FALSE) + stat_cor()
dev.off()

## Load parsimonious nodes
sample_data = read.delim("/groups/mpistaff/Zajac/genomes/sample_data.tsv")
pars_nodes = NULL
for (i in list.files("../pggb_allcommunities/", pattern = "parsimonious_dispnodes.gains.losses.txt", recursive = T)){
  dispnodes = read.delim(paste0("../pggb_allcommunities/", i))
  if (is.null(pars_nodes)){
    pars_nodes = dispnodes
  } else {
    pars_nodes = merge(pars_nodes, dispnodes[,c(4,10,11)], by = "label")
    pars_nodes$gains = pars_nodes$gains.x + pars_nodes$gains.y
    pars_nodes$losses = pars_nodes$losses.x + pars_nodes$losses.y
    pars_nodes = pars_nodes[,c(1:9,14,15)]
  }
}
tree = ape::read.tree("/groups/mpistaff/Zajac/genomes/pruned_tree.assembly_names.nwk")
p <- ggtree(tree) 
p$data = merge(p$data, pars_nodes, by = colnames(pars_nodes)[c(2:9)])
p$data = merge(sample_data[,c(3,17)], p$data, by.y = "label.x", by.x = "MyName", all.y = TRUE) %>% mutate(label = if_else(grepl("Node", label.y), label.y, Organism.Name)) %>% dplyr::select(3:10,12:14)
p$data = p$data %>% unique()

pdf("../Figures/Disp_nodes_phylogenetically_congruent_tree.pdf", width = 15, height = 11)
p +
    geom_tiplab(offset = 0.02, size = 5, fontface = "italic") +
    
    geom_point(data = p$data,
               aes(x = x, size = gains),
               shape = 21,
               fill = "darkgreen",
               color = "black",
               alpha = 0.85
    ) +
    
    geom_point( data = p$data,
                aes(x = x - dx, size = losses),
                shape = 21,
                fill = "lightgreen",
                color = "black",
                alpha = 0.85
    ) +
    
    geom_label(
        aes(
            label = paste0("+", gains, " / −", losses)
        ),
        size = 5,
        vjust = -0.8,
        fill = "beige"
    ) +
    
    theme_tree2() +
    xlim(0, 0.24) +
    scale_size_continuous(
        range = c(1.5, 8),
        name = "Number of events (Mio)"
    ) +
    theme(text = element_text(face = "italic"))
dev.off()
```

## 11. Dispensable nodes - how many shared between how many species

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}

genome_distribution_all = data.frame(nb_species = seq(2,14,1))
nb_nodes_only_one_species_missing_all = data.frame(rowname = sample_data$MyName[c(1:2,4,6:13,15:18)])
for (i in seq(0,18,1)[17:18]){
    files = list.files(paste0("../pggb_allcommunities/community", i, ".s5000/"), pattern = ".paths", full.names = T)[!grepl("#|all|sorted.paths", list.files(paste0("../pggb_allcommunities/community", i, ".s5000/"), pattern = ".paths"))]
  disp = list.files(paste0("../pggb_allcommunities/community", i, ".s5000/"), pattern = "dispensable", full.names = T)
  disp = fread(disp, header = F)
  
  dt_list <- lapply(files, function(i) {
    name <- str_split(i, "\\.")[[1]][6]
    x <- fread(i, header = FALSE)
    x[, (name) := 1]
    x[, .SD, .SDcols = c("V1", name)]
  })
  
  all_files <- rbindlist(dt_list, fill = TRUE)
  all_files <- all_files[, lapply(.SD, max, na.rm = TRUE), by = V1]
  
  disp <- merge(disp, all_files, by = "V1", all.x = TRUE)
  for (j in setdiff(names(disp), "V1")) {
    set(disp, which(is.na(disp[[j]])), j, 0)
  }
  disp = disp %>% column_to_rownames("V1")
  disp = disp %>% dplyr::rename("rPodCre2.1" = "rPodCre2") %>% dplyr::rename("rPodLil1.2" = "rPodLil1") 
  disp$nb_species = rowSums(disp)
  
  genome_distribution = disp %>% group_by(nb_species) %>% summarise(count = n())
  colnames(genome_distribution) = c("nb_species", paste0("community", i))
  genome_distribution_all = merge(genome_distribution_all, genome_distribution, by = "nb_species", all = TRUE)
  
  nb_nodes_only_one_species_missing = data.frame(value = colSums(disp[disp$nb_species == 14,] == 0)) %>% rownames_to_column() %>% filter(!rowname %in% c("nb_species"))
  colnames(nb_nodes_only_one_species_missing) = c("rowname", paste0("community", i))
  nb_nodes_only_one_species_missing_all = merge(nb_nodes_only_one_species_missing_all, nb_nodes_only_one_species_missing, by = "rowname", all = TRUE)
}

#Which species have the most disp nodes where it is missing
nb_nodes_only_one_species_missing_all %>% reshape2::melt(id.vars = "rowname") %>% dplyr::rename("community" = "variable", "MyName" = "rowname")  %>% left_join(chr_map) %>% left_join(sample_data[,c(3,16,17)]) %>% unique() %>% ggplot(aes(value,Organism.Name, fill = chromosome)) + geom_col() + theme_bw() + scale_fill_manual(values = brewPalette(n = 19)) + facet_grid(clade~., space = "free", scales = "free") + labs(x = "Dispensable nodes missing only in this species", y = "")

#Check how many nodes have how many species
genome_distribution_all %>% reshape2::melt(id.vars = "nb_species") %>% dplyr::rename("community" = "variable") %>% left_join(chr_map) %>% ggplot(aes(nb_species, value, fill = chromosome)) + geom_col() + scale_fill_manual(values = brewPalette(n=19)) + theme_bw() + labs(x = "Number of species", y  = "Number of dispensable nodes") + theme(text = element_text(size = 14))

#Compare the busco scores with number of disp nodes where only this species is missing
busco_results = read.delim("/groups/mpistaff/Zajac/analyses/BUSCO/summary.txt")
nb_nodes_only_one_species_missing_all %>% reshape2::melt(id.vars = "rowname") %>% dplyr::rename("community" = "variable", "MyName" = "rowname")  %>% left_join(chr_map) %>% left_join(sample_data[,c(3,16,17)]) %>% left_join(busco_results %>% separate(Genome, into = c("MyName", "haplotype"), sep = "#") %>% group_by(MyName) %>% summarise(Complete_Busco = mean(Complete_Busco))) %>% unique() %>% group_by(MyName, Organism.Name, clade) %>% summarise(Complete_Busco = mean(Complete_Busco), value = sum(value)) %>% ggplot(aes(Complete_Busco, value, colour = Organism.Name)) + geom_point(size = 4) + scale_colour_manual(values = brewPalette(15)) + theme_bw() + labs(y = "Number of dispensable nodes missing only in that species", x = "Complete BUSCO % (genomic)", colour = "Species Name") + theme(text = element_text(size = 14))

#Percentage of disp nodes per species where it is the only one missing
for (i in seq(0,18,1)) {value = as.numeric(system(paste("cat", paste0("../pggb_allcommunities/community", i, ".s5000/community", i ,".sorted.allSpecies.nodes.dispensable"), "| wc -l"), intern = TRUE)); comm = paste0("community",i); res = rbind(res, data.frame("community"= comm, value = value))}
merge(nb_nodes_only_one_species_missing_all %>% reshape2::melt(id.vars = "rowname") %>% dplyr::rename("community" = "variable", "MyName" = "rowname"), res, by = "community") %>% mutate(perc = (value.x  * 100) / value.y) %>% left_join(chr_map) %>% left_join(sample_data[,c(3,16,17)]) %>% unique() %>% ggplot(aes(perc,Organism.Name, colour = chromosome)) + geom_point() + theme_bw() + scale_colour_manual(values = brewPalette(n = 19)) + facet_grid(clade~., space = "free", scales = "free") + labs(x = "Percentage of dispensable nodes missing only in this species", y = "") + theme(text = element_text(size = 14))

#How many nodes each clade shares with Tiliguerta
df = merge(disp[disp$rPodTil1 == 1,c(1:15)] %>% rownames_to_column("node") %>% reshape2::melt(id.vars = "node") %>% filter(value == 1), sample_data[,c(16,17)], by.x = "variable", by.y = "MyName") 
df = df %>% unique()
df %>% group_by(clade) %>% summarise(n_distinct(node))

```

## 12. Order of disp/private/core nodes

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
#This is for one genome - this is run as a separate script 06_label_nodes_by_order_on_chr.R
# files = list.files("../pggb_allcommunities/", pattern = "*PodMur1#1#", recursive = TRUE)
# chr_map = read.delim("../chromosome_sets/community.chr.map")
# df_all_chr = NULL
# for (i in files){
#   comm = sapply(str_split(i, "/"), .subset, 1)
#   chr = chr_map[chr_map$community == str_remove(str_remove(comm, "community"), ".s5000"),]$chromosome
#   dispensable = list.files(paste0("../pggb_allcommunities/", comm), pattern = "nodes.dispensable", full.names = T)
#   dispensable = read.delim(dispensable, header = F)
#   private = list.files(paste0("../pggb_allcommunities/", comm), pattern = "nodes.private",full.names = T)
#   private = read.delim(private, header = F)
#   core = list.files(paste0("../pggb_allcommunities/", comm), pattern = "nodes.core",full.names = T)
#   core = read.delim(core, header = F)
#   len = list.files(paste0("../pggb_allcommunities/", comm), pattern = ".sorted.nodes.fasta.len",full.names = T)
#   len = read.delim(len, header = F)
#   df = read.delim(paste0("../pggb_allcommunities/", i), header = F)
#   colnames(df) = c("V1", "orientiation")
#   df = df %>% left_join(rbind(dispensable %>% mutate(cat = "dispensable"), private %>% mutate(cat = "private"), core %>% mutate(cat = "core"))) %>% left_join(len) 
#   df = df %>% mutate(block = cumsum(cat != lag(cat, default = first(cat)))) %>%
#     group_by(block, cat) %>%
#     summarise(V2 = sum(V2), segment_id = paste(V1, collapse = ","), .groups = "drop") %>% 
#     ungroup() %>% 
#     mutate(start = c(1, head(cumsum(V2), -1) + 1)) %>% mutate(end = cumsum(V2)) %>% 
#     mutate(chr = chr)
#   df_all_chr[[comm]] = df
# }

df = read.delim("../PhyloTree/GenomicMosaicism/community9.sorted.PodMur1#1#Z.paths.order.length.txt")

df %>% 
  ggplot(aes(x = start, xend = end, y = 1, yend = 1, colour = cat )) + 
  geom_segment(linewidth = 7) +
  scale_color_manual(
    values = c(
      private = "pink3",
      dispensable = "darkblue",
      core = "darkcyan"))
```

## 13. TEs in private/disp/core

```{r, fig.height=7, fig.width=15, echo=FALSE, warning=FALSE}
all_Tes = NULL
for (i in list.files("../Species_Specific_Masking/for_pggb/", pattern = "classification", full.names = T)){
  df = read.delim(i)
  df = df %>% separate(chr, into = c("species", "haplotype", "chr"), sep = "#")
  all_Tes = rbind(all_Tes, df)
}
all_Tes$TEclass =  str_remove(sapply(str_split(all_Tes$TE, "_"), .subset, 1), "\\?")
all_Tes = rbind(all_Tes %>% group_by(TEclass,cat,species,haplotype,chr) %>% dplyr::summarise(length = sum(length), total_length = mean(total_length)) %>% mutate(perc = length*100/total_length) %>% dplyr::select(c(1:5,8)), unique(all_Tes[,c(-5,-6,-9)]) %>% mutate(perc = exact_overlap*100/total_length) %>% dplyr::select(c(1:4,7)) %>% mutate(TEclass = "total"))
all_Tes = all_Tes %>% left_join(data.frame("TE_class" = c("ClassII: DNA Transposon", "ClassI: LINE", "ClassI: LTR", "ClassI: Low complexity", "Penelope-like", "ClassII: Rolling-circle", "ClassI: Retrotransposon", "ClassI: SINE", "Satellite repeats", "Simple repeats", "rRNA", "snRNA", "tRNA", "ALL"), "TEclass" = unique(all_Tes$TEclass)))
all_Tes$TE_class = factor(all_Tes$TE_class, levels = unique(all_Tes$TE_class))
pdf("../Figures/PG_TEandREpeats.pdf", width = 9, height = 5)
all_Tes %>% mutate(pang = if_else(TE_class == "ALL", "ALL", "Repeats/TEs")) %>% ggplot(aes(TE_class, perc, fill = cat)) + geom_boxplot(outliers = F) + theme_bw() + labs(x = "", y = "Repeat density (%)", fill = "Pangenome Class") + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), text = element_text(size = 15))  + scale_fill_manual(values = c("#0072b2", "orange","#cc79a7")) + facet_grid(~pang, scales = "free", space = "free") 
dev.off()
```
