---
title: "SVs in TEs"
author: "Natalia Zajac"
date: "05/11/2025"
output:
  html_document: 
    highlight: pygments
    theme: sand
    code_folding: true
    toc: true
    toc_float: true
---

# This notebook aims to evaluate and analyse SV data

```{r setup, echo=FALSE, include=FALSE, warning=FALSE}
library(vcfR)
library(utils)
library(dplyr)
library(tidyverse)
library(reshape2)
library(cowplot)
library(patchwork)
library(ezRun)
library(ggbeeswarm)
library(data.table)
library(DT)
library(parallel)

# Metadata
sample_data = read.delim("/groups/mpistaff/Zajac/genomes/sample_data.tsv")
sample_data$clade=factor(sample_data$clade,levels=c("Balkan","Sicilian-Maltese","Western","Siculus","Muralis","Iberian"))


# Bed coordinates for cretensis
bed = read.delim("/groups/mpistaff/Zajac/genomes/ncbi_dataset/data/ALL_corrected_unmasked/rPodCre2.1#1.all.fasta.fai", header = F)
bed$Start = 1
bed = bed[,c(1,6,2)]
colnames(bed) = c("Chr", "Start", "End")

#Load bed files for all species
beds = list.files("/groups/mpistaff/Zajac/genomes/ncbi_dataset/data/ALL_corrected_unmasked/", pattern = "fai", full.names = TRUE)
beds = lapply(beds, function(x){ 
  bed = read.delim(x, header = F)
  bed$Start = 1
  bed = bed[,c(1,6,2)]
  colnames(bed) = c("Chr", "Start", "End")
  bed  = bed %>% separate(Chr, into = c("Focal_genome",  "Focal_hap", "Focal_chr"), sep = "#")
  return(bed)
})
beds = bind_rows(beds)

# Repeats and TEs - classification
summary_repeats =  fread("../Species_Specific_Masking/repeats.allhaplotypes.summary.tsv", sep = "\t")
summary_repeats = summary_repeats[,c(1:9)]
colnames(summary_repeats) = c("Focal_genome", "Focal_hap", "Focal_chr", colnames(summary_repeats)[-c(1:3)])
summary_repeats$Focal_hap = as.character(summary_repeats$Focal_hap)
summary_repeats = summary_repeats %>% 
  left_join(beds, by = c("Focal_genome", "Focal_hap", "Focal_chr")) %>% 
  mutate(blocks = round(End/16,0))
summary_repeats = summary_repeats %>%
  group_by(Focal_genome, Focal_hap, Focal_chr) %>%
  group_modify(~ {
    breaks <- seq(unique(.x$blocks), unique(.x$End) - unique(.x$blocks), length.out = 15)
    .x %>%
      mutate(block = ezCut(end, 
                           breaks, labels = as.character(seq(1,16,1))))
  }) %>%
  ungroup()
summary_repeats = summary_repeats %>% 
  group_by(Focal_genome, Focal_hap, Focal_chr, block, blocks) %>% 
  summarise(repeat_covered_bases = sum(repeat_covered_bases)) %>% 
  mutate(repeat_covered_bases = repeat_covered_bases*100/as.numeric(blocks))
summary_repeats = summary_repeats[summary_repeats$Focal_chr != "W",]
summary_repeats$chr_category = case_when(summary_repeats$Focal_chr  %in% c("Z") ~ "Z", summary_repeats$Focal_chr %in% c("18") ~ "18", .default = "17 Autosomes")

repeats =  fread("../Species_Specific_Masking/repeats.allhaplotypes.tsv", sep = "\t")
repeats$seqid = paste0(repeats$species, "#", repeats$haplotype, "#", repeats$chromosome)
repeats_gr <- makeGRangesFromDataFrame(repeats[,c(4,5,6,13)], 
                                       seqnames.field = "seqid", 
                                       start.field = "positioninquery_begin", 
                                       end.field = "positioninquery_end", 
                                       keep.extra.columns = T)

# SVs (INDELs)
all_comb = NULL
for (i in list.files("../SV_intersects/largeSVs/", pattern = "merged.DEFAULT")[!grepl("annot",list.files("../SV_intersects/largeSVs/", pattern = "merged.DEFAULT"))]){
    comb = read.vcfR(paste0("../SV_intersects/largeSVs/", i))
    comb = cbind(data.frame(comb@fix), data.frame(comb@gt))
    comb = comb %>% 
      mutate(countALT = nchar(ALT), countN = str_count(ALT, "N")) %>% 
      mutate(perc = countN/countALT) %>% filter(perc < 0.5) 
    comb = comb %>% dplyr::mutate(across(10:23, ~ str_remove(., ":.*"))) %>% 
        mutate(REFLEN = nchar(REF), ALTLEN = nchar(ALT)) %>% 
        mutate(SVTYPE = sub(".*SVTYPE=([^;]+).*", "\\1", INFO),
               TYPE=sub(".*;TYPE=([^;]+).*", "\\1", INFO),
               AVG_LEN = sub(".*AVG_LEN=([^;]+).*", "\\1", INFO), 
               AVG_START = sub(".*AVG_START=([^;]+).*", "\\1", INFO), 
               AVG_END = sub(".*AVG_END=([^;]+).*", "\\1", INFO)) %>% 
        mutate(SVTYPE = case_when(SVTYPE == "INS" ~ "INS", 
               SVTYPE == "DEL" ~ "DEL", 
               (!SVTYPE %in% c("INS", "DEL")) &  !is.na(TYPE) ~ TYPE, 
               (!SVTYPE %in% c("INS", "DEL")) &  is.na(TYPE) ~ "NA"))
    comb = comb[,c(1,2,3,10:23,27:33)]
    comb = comb %>% dplyr::mutate(across(colnames(comb)[4:17], ~ case_when(. == "0|0" ~ 0, . == "0" ~ 0, . == "1|1" ~ 1, . == "1" ~ 1,  . == "0|1" ~ 1, . == "1|0" ~ 1, . == ".|1" ~ 1, . == "0|."~ 0, . == ".|0" ~ 0, . == "1|." ~ 1 )))
    all_comb[[i]] = comb
}
all_comb = bind_rows(all_comb)
all_comb$LEN = abs(all_comb$ALTLEN - all_comb$REFLEN)
all_comb$nb_species = rowSums(all_comb[,c(4:17)] > 0)
for_karyotype = all_comb[all_comb$nb_species >=1,]
kary = NULL
for (i in unique(for_karyotype$CHROM)){
  tmp = for_karyotype[for_karyotype$CHROM ==i,]
  bed1 = bed[bed$Chr ==i,]$End
  tmp$blocks = ezCut(x = as.numeric(tmp$POS), breaks = seq(1,as.numeric(bed1), 500000))
  kary = rbind(kary,tmp)
}
kary$POS = as.numeric(kary$POS)
kary = kary %>%
    group_by(CHROM) %>%
    mutate(
        max_pos = max(POS, na.rm = TRUE),
        blocks = if_else(
            str_detect(blocks, "^>\\s*\\d+"),
            str_replace(blocks, "^>\\s*(\\d+)", "(\\1 - ") %>%
                paste0(max_pos, "]"),
            blocks
        )
    ) %>%
    select(-max_pos)
kary = kary[,c(1,2,27)] %>% group_by(blocks, CHROM) %>% summarise(n_distinct(POS)) %>% mutate(Start =as.numeric(sapply(str_split(str_remove(as.character(blocks), "\\("), " - "), .subset, 1)), End = as.numeric(sapply(str_split(str_remove(as.character(blocks), "\\]"), " - "), .subset, 2))) %>% ungroup() %>% dplyr::select(c(2,4,5,3))
colnames(kary) = c("Chr", "Start", "End", "Value")
kary = merge(kary, bed, by = "Chr") %>%  
    mutate(blocks = round(End.y/16,0)) %>%
    group_by(Chr) %>%
    group_modify(~ {
        breaks <- seq(unique(.x$blocks), unique(.x$End.y) - unique(.x$blocks), length.out = 15)
        .x %>%
            mutate(block = ezCut(End.x, 
                                 breaks, labels = as.character(seq(1,16,1))))
    }) %>%
    ungroup() %>% 
    group_by(Chr, block, blocks) %>% 
    summarise(SVdensity = sum(Value)) 
kary$chr_category = case_when(kary$Chr  %in% c("rPodCre2.1#1#Z") ~ "Z", kary$Chr %in% c("rPodCre2.1#1#18") ~ "18", .default = "17 Autosomes")

#Load SNPs and block
snps = readRDS("../pggb_allcommunities/general_stats/SNP.rideogram.data.rds")
snps = merge(snps, bed, by.x = "CHROM", by.y = "Chr") %>%  
    mutate(blocks = round(End/16,0)) %>%
    group_by(CHROM) %>%
    group_modify(~ {
        breaks <- seq(unique(.x$blocks), unique(.x$End) - unique(.x$blocks), length.out = 15)
        .x %>%
            mutate(block = ezCut(end, 
                                 breaks, labels = as.character(seq(1,16,1))))
    }) %>%
    ungroup() %>% 
    group_by(CHROM, block, blocks) %>% 
    summarise(SVdensity = sum(value)) 
snps$chr_category = case_when(snps$CHROM  %in% c("rPodCre2.1#1#Z") ~ "Z", snps$CHROM %in% c("rPodCre2.1#1#18") ~ "18", .default = "17 Autosomes")

snps %>% ggplot(aes(block, SVdensity)) + geom_violin(fill = "#3C561E") + theme_bw() + labs(x = "genomic blocks along chromosomes", y = "SNPs Count") + theme(text = element_text(size = 15), axis.text.x = element_blank()) + facet_grid(chr_category~., scales = "free")

#Dispensable nodes legnth of nodes summary
dispfiles = list.files("../pggb_allcommunities/", pattern = "nodes.fasta.len", full.names = T, recursive = T)
disppg = NULL
for (i in dispfiles){
  df = fread(i)
  df$blocks = ezCut(df$V2,seq(1,100,49))
  disppg = rbind(disppg, df %>% group_by(blocks) %>% summarise(sum(V2), count = n()))
}

#Dispensable nodes distibution
dispnodes = read.delim("/home/zajac/PhyloTree/GenomicMosaicism/all_chr_node_cats.txt", sep = "\t")
dispnodes$chr_category = case_when(dispnodes$Focal_chr  %in% c("Z") ~ "Z", dispnodes$Focal_chr %in% c("18") ~ "18", .default = "17 Autosomes")


#Load INVs
all_invs = read.delim("../INVERSIONS/inversions.syri.impg.mapc.filtered.clustered.0.9.txt")
shared = all_invs[all_invs$is_singleton == "FALSE",] %>% 
    dplyr::select("species","chr", "max_start" ,"min_end", "mean_length") %>% 
    separate(species, into = c("species", "haplotype", "chr"), sep = "#") %>%
    dplyr::select(1,3,4,5,6) %>% 
    unique() %>% 
    pivot_wider(names_from = species, values_from = mean_length) %>% 
    replace(is.na(.),0) %>% mutate(across(c(4:17), ~ if_else(. > 0, 1,0))) %>%
    mutate(N = rowSums(.[4:17])) 
all_invs$shared_between_species = ""
all_invs[paste0(all_invs$max_start, "-", all_invs$min_end) %in% paste0(shared[shared$N >1,]$max_start,"-",shared[shared$N >1,]$min_end),]$shared_between_species = "Yes"

# Check overlap with repeats
all_invs_gr <- makeGRangesFromDataFrame(all_invs[,c("species", "start_species_syri", "end_species_syri", "shared_between_species")], 
                                     seqnames.field  =  "species", start.field = "start_species_syri", end.field = "end_species_syri", keep.extra.columns = T
)
hits = findOverlaps(all_invs_gr, repeats_gr)
query_hits   <- all_invs_gr[queryHits(hits)]
subject_hits <- repeats_gr[subjectHits(hits)]
overlaps <- pintersect(query_hits, subject_hits)
result <- data.frame(
    query_seqname   = as.character(seqnames(query_hits)),
    query_start     = start(query_hits),
    query_end       = end(query_hits),
    inv_len         = end(query_hits) - start(query_hits),
    shared          = query_hits$shared_between_species,
    subject_seqname = as.character(seqnames(subject_hits)),
    subject_start   = start(subject_hits),
    subject_end     = end(subject_hits),
    repeat_len      = end(subject_hits) - start(subject_hits),
    overlap_start   = start(overlaps),
    overlap_end     = end(overlaps),
    repeats = subject_hits$repeat_class,
    overlap_length  = width(overlaps)
) %>% unique()
inv_overlap = result %>% group_by(query_seqname, query_start, query_end, shared) %>% dplyr::summarise(overlap_length = sum(overlap_length), repeat_len = sum(repeat_len), inv_len = mean(inv_len)) 

#Load overlap of INDELs with TEs
insertions = fread("../Species_Specific_Masking/transposable_elements_fasta/temp.INS.filt.txt")
deletions = fread("../Species_Specific_Masking/transposable_elements_fasta/temp.DEL.filt.txt")
```


# Proportion of different size blocks in pangenome

```{r}
p1 = disppg %>% group_by(blocks) %>% summarise(length = sum(`sum(V2)`), count = sum(count)) %>% ggplot(aes(factor(blocks, levels = c("<= 1", "(1 - 50]", "> 50")),count/1000000)) + geom_col() + theme_bw() + labs(x = "Node length (bp)", y = "") + ggtitle("Count (Mio)")
p2 = disppg %>% group_by(blocks) %>% summarise(length = sum(`sum(V2)`), count = sum(count)) %>% ggplot(aes(factor(blocks, levels = c("<= 1", "(1 - 50]", "> 50")),length/1000000000)) + geom_col() + theme_bw() + labs(x = "Node length (bp)", y = "") + ggtitle("Length (Gbp)")

p1 + p2

pdf("../Figures/Disp_node_length.pdf", width = 5, height = 5)
p1 + p2
dev.off()
```

# Combined distribution of SNPs, Dispnodes, SVs and Repeats

```{r}

dispnodes$block_factor <- factor(
    dispnodes$block,
    levels = 1:16
)

summary_repeats$block_factor <- factor(
    summary_repeats$block,
    levels = 1:16
)

kary$block_factor <- factor(
    kary$block,
    levels = 1:16
)

snps$block_factor <- factor(
    snps$block,
    levels = 1:16
)

# Dispensable nodes
p3 <- ggplot(
    dispnodes[dispnodes$cat == "dispensable", ],
    aes(block_factor, value)
) +
    geom_violin(
        fill = "grey75",
        colour = "grey30",
        linewidth = 0.25,
        width = 0.9,
        trim = FALSE
    ) +
    facet_grid(
        . ~ chr_category,
        scales = "free_y",
        space = "fixed"
    ) +
    scale_y_continuous(
        n.breaks = 4,
        expand = expansion(mult = c(0.05, 0.05))
    ) +
    labs(
        y = "Dispensable nodes (%)"
    ) +
    theme_classic(base_size = 10) +
    theme(
        # Remove x-axis
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        
        # Show chromosome labels only here
        strip.background = element_blank(),
        strip.text = element_text(
            size = 9,
            face = "bold"
        ),
        
        panel.spacing = unit(0.25, "lines"),
        
        axis.title.y = element_text(size = 9),
        axis.text.y = element_text(size = 8),
        
        plot.margin = margin(2, 3, 0, 3)
    )


# TEs and repeats

p4 <- ggplot(
    summary_repeats,
    aes(block_factor, repeat_covered_bases)
) +
    geom_violin(
        fill = "grey75",
        colour = "grey30",
        linewidth = 0.25,
        width = 0.9,
        trim = FALSE
    ) +
    facet_grid(
        . ~ chr_category,
        scales = "free_y",
        space = "fixed"
    ) +
    scale_y_continuous(
        n.breaks = 4,
        expand = expansion(mult = c(0.05, 0.05))
    ) +
    labs(
        y = "TEs & repeats (%)"
    ) +
    theme_classic(base_size = 10) +
    theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        
        # Hide repeated chromosome labels
        strip.background = element_blank(),
        strip.text = element_blank(),
        
        panel.spacing = unit(0.25, "lines"),
        
        axis.title.y = element_text(size = 9),
        axis.text.y = element_text(size = 8),
        
        plot.margin = margin(0, 3, 0, 3)
    )


#Indels 

p5 <- ggplot(
    kary,
    aes(
        block_factor,
        SVdensity / 1000
    )
) +
geom_violin(
    data = subset(
        kary,
        chr_category == "17 Autosomes"
    ),
    fill = "grey75",
    colour = "grey30",
    linewidth = 0.25,
    width = 0.9,
    trim = FALSE
) +
geom_point(
    data = subset(
        kary,
        chr_category %in% c("18", "Z")
    ),
    position = position_jitter(
        width = 0.10,
        height = 0,
        seed = 1
    ),
    size = 1.5,
    alpha = 0.7
) +
    
    facet_grid(
        . ~ chr_category,
        scales = "free_y",
        space = "fixed"
    ) +
    
    scale_y_continuous(
        n.breaks = 4,
        expand = expansion(mult = c(0.05, 0.05))
    ) +
    
    labs(
        y = "INDELs (≥50 bp) (×10³)"
    ) +
    
    theme_classic(base_size = 10) +
    theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        
        # Chromosome labels already shown at top
        strip.background = element_blank(),
        strip.text = element_blank(),
        
        panel.spacing = unit(0.25, "lines"),
        
        axis.title.y = element_text(size = 9),
        axis.text.y = element_text(size = 8),
        
        plot.margin = margin(0, 3, 0, 3)
    )


#SNPs

p6 <- ggplot(
    snps,
    aes(
        block_factor,
        SVdensity / 10000
    )
) +
    
geom_violin(
    data = subset(
        snps,
        chr_category == "17 Autosomes"
    ),
    fill = "grey75",
    colour = "grey30",
    linewidth = 0.25,
    width = 0.9,
    trim = FALSE
) +
    
geom_point(
    data = subset(
        snps,
        chr_category %in% c("18", "Z")
    ),
    position = position_jitter(
        width = 0.10,
        height = 0,
        seed = 1
    ),
    size = 1.5,
    alpha = 0.7
) +
    
    facet_grid(
        . ~ chr_category,
        scales = "free_y",
        space = "fixed"
    ) +
    
    scale_y_continuous(
        n.breaks = 4,
        expand = expansion(mult = c(0.05, 0.05))
    ) +
    
    labs(
        y = "SNPs (×10⁴)",
        x = "Genomic blocks along chromosomes"
    ) +
    
    theme_classic(base_size = 10) +
    theme(
        # Only bottom plot gets x-axis
        axis.title.x = element_text(size = 9),
        axis.text.x = element_text(size = 7),
        axis.ticks.x = element_line(linewidth = 0.25),
        
        # Chromosome labels already shown at top
        strip.background = element_blank(),
        strip.text = element_blank(),
        
        panel.spacing = unit(0.25, "lines"),
        
        axis.title.y = element_text(size = 9),
        axis.text.y = element_text(size = 8),
        
        plot.margin = margin(0, 3, 2, 3)
    )


# Combine 
final_plot <- p3 / p4 / p5 / p6 +
    plot_layout(
        guides = "collect",
        heights = c(1, 1, 1, 1)
    )

final_plot


pdf("../Figures/Distribution_all_combined.pdf", width = 10, height = 8)
final_plot
dev.off()
```

# Percentage of TEs in different genomic features


```{r}
deletions$perc_overlap = deletions$overlap_length * 100/deletions$width
insertions$perc_overlap = insertions$leninTE * 100/insertions$LEN
inv_overlap$perc_overlap = inv_overlap$overlap_length*100/inv_overlap$inv_len

p7 = rbind(inv_overlap[,c("query_seqname", "perc_overlap", "shared")] %>% 
             dplyr::rename("seqnames" = "query_seqname") %>% 
             mutate("Feature" = if_else(shared == "Yes", "shared INV", "unique INV"), seqnames = paste0(sapply(str_split(seqnames, "#"), .subset, 1), "#", sapply(str_split(seqnames, "#"), .subset, 2)))  %>% 
             group_by(Feature, seqnames) %>% 
             dplyr::summarise(perc_overlap = mean(perc_overlap)),
      summary_repeats %>% 
        group_by(Focal_genome, Focal_hap) %>% 
        summarise(perc_overlap = mean(repeat_covered_bases)) %>% 
        mutate("seqnames" = paste0(Focal_genome, "#", Focal_hap)) %>% 
        ungroup() %>% dplyr::select(4,3) %>% mutate("Feature" = "Genome-wide"),
      deletions[,c("seqnames","perc_overlap")] %>% 
        mutate("Feature" = "DEL (>=50bp)") %>% 
        mutate(perc_overlap = if_else(perc_overlap > 100, 100, perc_overlap)) %>% 
        group_by(Feature) %>% dplyr::summarise(perc_overlap = mean(perc_overlap)),
      insertions[,c("species", "perc_overlap")] %>% 
        dplyr::rename("seqnames" = "species") %>% 
        mutate("Feature" = "INS (>=50bp)") %>% 
        mutate(perc_overlap = if_else(perc_overlap > 100, 100, perc_overlap)) %>% 
        group_by(Feature, seqnames) %>% 
        dplyr::summarise(perc_overlap = mean(perc_overlap))
) %>% 
  ggplot(aes(factor(Feature, levels = c("Genome-wide", "DEL (>=50bp)", "INS (>=50bp)", "shared INV", "unique INV")),perc_overlap )) + 
  geom_boxplot() + theme_bw() + labs(x = "", y = "% of length in TEs and Repeats") + 

p7 

pdf("../Figures/Repeats_in_features.pdf", width = 5, height = 5)
p7 +  stat_compare_means(comparisons = combn(unique(p7$data$Feature), 2, simplify = FALSE)[c(2,5)], method = "wilcox.test", label = "p.signif", hide.ns = TRUE)
dev.off()

```

# Percentage of TEs in different genomic features


```{r}
deletions$perc_overlap = deletions$overlap_length * 100/deletions$width
insertions$perc_overlap = insertions$leninTE * 100/insertions$LEN
inv_overlap$perc_overlap = inv_overlap$overlap_length*100/inv_overlap$inv_len

p7 = rbind(inv_overlap[,c("query_seqname", "perc_overlap", "shared")] %>% 
             dplyr::rename("seqnames" = "query_seqname") %>% 
             mutate("Feature" = if_else(shared == "Yes", "shared INV", "unique INV"), seqnames = paste0(sapply(str_split(seqnames, "#"), .subset, 1), "#", sapply(str_split(seqnames, "#"), .subset, 2)))  %>% 
             group_by(Feature, seqnames) %>% 
             dplyr::summarise(perc_overlap = mean(perc_overlap)),
      summary_repeats %>% 
        group_by(Focal_genome, Focal_hap) %>% 
        summarise(perc_overlap = mean(repeat_covered_bases)) %>% 
        mutate("seqnames" = paste0(Focal_genome, "#", Focal_hap)) %>% 
        ungroup() %>% dplyr::select(4,3) %>% mutate("Feature" = "Genome-wide"),
      deletions[,c("seqnames","perc_overlap")] %>% 
        mutate("Feature" = "DEL (>=50bp)") %>% 
        mutate(perc_overlap = if_else(perc_overlap > 100, 100, perc_overlap)) %>% 
        group_by(Feature) %>% dplyr::summarise(perc_overlap = mean(perc_overlap)),
      insertions[,c("species", "perc_overlap")] %>% 
        dplyr::rename("seqnames" = "species") %>% 
        mutate("Feature" = "INS (>=50bp)") %>% 
        mutate(perc_overlap = if_else(perc_overlap > 100, 100, perc_overlap)) %>% 
        group_by(Feature, seqnames) %>% 
        dplyr::summarise(perc_overlap = mean(perc_overlap))
) %>% 
  ggplot(aes(factor(Feature, levels = c("Genome-wide", "DEL (>=50bp)", "INS (>=50bp)", "shared INV", "unique INV")),perc_overlap )) + 
  geom_boxplot() + theme_bw() + labs(x = "", y = "% of length in TEs and Repeats") 

p7 

pdf("../Figures/Repeats_in_features.pdf", width = 5, height = 5)
p7 +  stat_compare_means(comparisons = combn(unique(p7$data$Feature), 2, simplify = FALSE)[c(2,5)], method = "wilcox.test", label = "p.signif", hide.ns = TRUE)
dev.off()

